---
title: "Cyclistic Case Study"
author: "Oluwatunmise Ayinde"
date: '2022-05-17'
output:
  html_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: inline
---

## Background for this Activity
In this activity, I will review a scenario, and focus on manipulating and changing real data in R. I will learn more about functions I can use to manipulate my data, use statistical summaries to explore my data, and gain initial insights for stakeholders. 

Throughout this activity, i will also have the opportunity to practice writing my own code by making changes to the code chunks yourself. 

## The Scenario
In 2016, Cyclistic launched a successful bike-share offering. Since then, the program has grown to a fleet of 5,824 bicycles that are geotracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and returned to any other station in the system anytime.

There are two types of cyclists, those who purchase casual tickets and those purchase annual memberships. The marketing team believes that maximizing the number of annual members will be key to future growth. Rather than creating a marketing campaign that targets all-new-customers, there is a very good chance to convert casual riders into members.

## Problem
What is the most effective marketing strategy of converting Cyclistic’s casual riders to annual memberships?
* How do annual members and casual riders use Cyclistic’s bikes differently?

## NOTE
This project was initially started with excel. Excel was used to clean the data, and to calculate the ride length of each trip and also to get the weekdays each trip started. Duplicate columns were removed, white spaces trimmed and all other basic data cleaning processes.

## Step 1: LOAD PACKAGES
```{r}
install.packages("tidyverse")
install.packages("janitor")
install.packages("skimr")
install.packages("lubridate")


library(tidyverse)
library(janitor)
library(skimr)
library(lubridate)
```

## Step 2: READ ALL CSV FILES
```{r}
df1 <-read_csv("202101-divvy-tripdata.csv")
df2 <-read_csv("202102-divvy-tripdata.csv")
df3 <-read_csv("202103-divvy-tripdata.csv")
df4 <-read_csv("202104-divvy-tripdata.csv")
df5 <-read_csv("202105-divvy-tripdata.csv")
df6 <-read_csv("202106-divvy-tripdata.csv")
df7 <-read_csv("202107-divvy-tripdata.csv")
df8 <-read_csv("202108-divvy-tripdata.csv")
df9 <-read_csv("202109-divvy-tripdata.csv")
df10 <-read_csv("202110-divvy-tripdata.csv")
df11 <-read_csv("202111-divvy-tripdata.csv")
df12 <-read_csv("202112-divvy-tripdata.csv")

View(df1)
```

## Step 3: PREVIEW THE DATA STRUCTURE OF THE MONTHLY DATA
This is done using the "str()" fuction
```{r}
str(df1)
str(df2)
str(df3)
str(df4)
str(df5)
str(df6)
str(df7)
str(df8)
str(df9)
str(df10)
str(df11)
str(df12)
```

## Step 4: CREATE THE DATAFRAME FOR ANALYSIS
Now its time to merge all the monthly data into one data frame so as to get a dataframe that contains a years worth analysis of the bike company. The "bind_rows()" function is used in this situation.
```{r}
bikes_rides <-bind_rows(df1,df2,df3,df4,df5,df6,df7,df8,df9,df10,df11,df12)

bikes_rides
```


# Starting the Analysis (cleaning)
## Step 5: DATA CLEANING

```{r}
# Rename columns to make them consistent

bikes_rides <-rename(bikes_rides, trip_id =ride_id, bike_type=rideable_type,start_time=started_at, end_time=ended_at,
                     from_station_name=start_station_name, from_station_id=start_station_id,to_station_name=end_station_name,
                     to_station_id=end_station_id,user_type=member_casual)

#cleaning the data

bikes_rides <- clean_names(bikes_rides)  #clean up non consistent variable names

bikes_rides <- remove_empty(bikes_rides, c("rows","cols"))  #remove empty rows and columns

bikes_rides <- subset(bikes_rides, select=-c(start_lat,start_lng,end_lat,end_lng)) #removing columns that are not required for the analysis

```

## Summary Statistics
The "nrow()" function is used to display the number of observations in the data frame.
The "head()" function is used to display the first 5 observations in the data frame
```{r}
nrow(bikes_rides)  

head(bikes_rides) 

summary(bikes_rides)
```


#Analyse Phase
## Descriptive analysis
```{r}
#Analyse phase
all_trips <- filter(bikes_rides, ride_length>0)
View(all_trips)


#aggregate our data based on the user_type column and find summary statistics
all_trips %>% 
  group_by(user_type) %>% 
  drop_na() %>% 
  summarise(mean_ride_length=mean(ride_length), min_ride_length=min(ride_length), max_ride_length=max(ride_length))

# find the day of week that rides are mostly used
getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

mode_day_of_week <- getmode(all_trips$day_of_week)
mode_day_of_week

#find the day of week thats most common based on user_type
all_trips %>% 
  group_by(user_type) %>% 
  drop_na() %>% 
  summarise(getmode(day_of_week))
```

## Analysis with Viz
```{r}
#find average ride_length per user_type per day of week and visualize
all_trips %>% 
  group_by(user_type, day_of_week) %>% 
  drop_na() %>% 
  summarise(average_ride=mean(ride_length)) %>% 
  ggplot() + geom_col(aes(x = day_of_week, y = average_ride, fill = user_type), position = "dodge") +
  scale_x_continuous(breaks = seq(0, 7, 1), 
                     limits=c(0, 8)) +
  labs(title = "AVERAGE RIDE LENGTH PER USER_TYPE DAILY", x = "Average ride length", y = "Day of week")+
  ggsave("AVERAGE RIDE LENGTH PER USER_TYPE DAILY.png")

```

```{r}
#find the average ride per user_type
all_trips %>% 
 group_by(user_type) %>% 
  drop_na() %>% 
  summarise(average_ride=mean(ride_length)) %>% 
  ggplot()+ geom_col(aes(x=user_type, y= average_ride, fill=user_type))+
  labs(title = "AVERAGE RIDE LENGTH PER USER_TYPE ", x = "User Type", y = "Average ride")+
  ggsave("AVERAGE RIDE LENGTH PER USER_TYPE.png")

```

```{r}
#find the number of rides per user_type daily
ride_per_day <-all_trips %>% 
  count(day_of_week, user_type)
ride_per_day   
  
ggplot(data = ride_per_day)+ geom_col(mapping = aes(x=day_of_week, y=n, fill=user_type), position = "dodge")+
  scale_x_continuous(breaks = seq(1, 7, 1), 
                     limits=c(0, 8))+
  labs(title = "Number of Rides Daily per User Type", x = "Day of week", y = "Number of Ride")+
  ggsave("Number of Rides Daily per User Type.png")

```

```{r}
#find the types of bikes mostly preferred by differnt user_type
taste <-all_trips %>% 
  count(bike_type, user_type)

taste
ggplot(data = taste)+ geom_col(mapping = aes(x=bike_type, y=n, fill=user_type), position = "dodge")+
  labs(title = "User Bike Preference", x = "Bike Type", y = "Number of User")+
  ggsave("User Bike Preference.png")
```

